Post MCMCglmm analyses

### 2021-04-26 Analysis/06-post_mcmcglmm_analyses.Rmd
Update 06-post_mcmcglmm_analyses.Rmd

```{r}
source("../Functions/get.axes.R")
## Get all the phylo main axes
level_centred_major_axes <- get.axes(covar_matrices, centre = "level")
uncentered_major_axes <- get.axes(covar_matrices, centre = "none")
```

@@

```{r}
plot(morphdat[, c(1,2)], pch = 19, cex = 0.5,
     col = colour_vector[morphdat$clade],
     xlab = "PC1 (90.5%)",
     ylab = "PC1 (6.86%)")
plot.ellipses(covar_matrices, centre = "none",
              col = c("grey",colour_vector),
              add = TRUE)
plot.all.axes(uncentered_major_axes,
         col = c("grey",colour_vector),
         add = TRUE)
```

@@

```{r}
plot(morphdat[, c(1,2)], pch = 19, cex = 0.5,
     col = colour_vector[morphdat$clade],
     xlab = "PC1 (90.5%)",
     ylab = "PC1 (6.86%)")
plot.ellipses(covar_matrices, centre = "level",
              col = c("grey",colour_vector),
              add = TRUE)
plot.all.axes(level_centred_major_axes,
         col = c("grey",colour_vector),
         add = TRUE)
```

@@

```{r}
plot(morphdat[, c(1,2)], pch = 19, cex = 0.5,
     col = colour_vector[morphdat$clade],
     xlab = "PC1 (90.5%)",
     ylab = "PC1 (6.86%)",
     main = "Major axes of the ellipses\n(aspect ratio = 1)",xlim = c(-1.5, 1.5), ylim = c(-1.5, 1.5))
plot.ellipses(covar_matrices, centre = "level",
              col = c("grey",colour_vector),
              add = TRUE)
plot.all.axes(level_centred_major_axes,
         col = c("grey",colour_vector),
         add = TRUE)
```

@@

```{r}
plot.all.axes(uncentered_major_axes,
         col = c("grey",colour_vector),
         add = FALSE, main = "Major axes (un-centred)")
```

@@

```{r}
source("../Functions/vector.diff.R")
## Get all the angles and stuff
gull_results <- mapply(vector.diff, uncentered_major_axes$cladegulls, uncentered_major_axes$phylo)
plovers_results <- mapply(vector.diff, uncentered_major_axes$cladeplovers, uncentered_major_axes$phylo)
sandpipers_results <- mapply(vector.diff, uncentered_major_axes$cladesandpipers, uncentered_major_axes$phylo)

@@

```{r, fig.height = 18, fig.heigh = 6}
## Plot the blob wise metrics
par(mfrow = c(3, 1))
boxplot(angles, col = colour_vector, ylab = "angle")
boxplot(projections, col = colour_vector, ylab = "projection (scaled)")
boxplot(rejections, col = colour_vector, ylab = "rejection (scaled)")
```

@@

```{r, fig.height = 18, fig.heigh = 6}
## Plot the blob wise metrics
par(mfrow = c(3, 1))
boxplot(angles %% 90, col = colour_vector, ylab = "angle (modulo 90)")
boxplot(abs(projections), col = colour_vector, ylab = "projection (scaled - absolute)")
boxplot(abs(rejections), col = colour_vector, ylab = "rejection (scaled - absolute)")
```

### Phylo clade model

For this example, the MCMCglmm model has four random (phylo) effects (one for the whole phylo and one per clade) rather than just the one for the whole phylogeny.
The model seems to struggle converging and there are things off with the centring (they seem to be all centred no matter what) but that last bit should not be a problem for comparing the vectors since they are translated anyways.

```{r, warn = FALSE}
plot(morphdat[, c(1,2)], pch = 19, cex = 0.5,
     col = colour_vector[morphdat$clade],
     xlab = "PC1 (90.5%)",
     ylab = "PC1 (6.86%)")
plot.ellipses(model_phylo4, n = 100,
              centre = "none", add = TRUE, col = c("grey",colour_vector, "pink"))
```

The pink ellipse is representing the residuals.

And we can still do the whole axis stuff, same way as above:

```{r}
## Get the covariance matrices
covar_matrices2 <- get.covar(model_phylo4,
                            n = 100,
                            simplify = FALSE)
## Drop the residuals
covar_matrices2 <- covar_matrices2[-5, ]

## Get all the phylo main axes
uncentered_major_axes2 <- get.axes(covar_matrices2, centre = "none")
names(uncentered_major_axes2) <- c("phylo", "cladegulls", "cladeplovers", "cladesandpipers")

```

Here's what the axes look like without the ellipses and stuff:

```{r}
plot.all.axes(uncentered_major_axes2,
         col = c("grey",colour_vector),
         add = FALSE, main = "Major axes (centred)")
```

```{r}
source("../Functions/vector.diff.R")
## Get all the angles and stuff
gull_results <- mapply(vector.diff, uncentered_major_axes2$cladegulls, uncentered_major_axes2$phylo)
plovers_results <- mapply(vector.diff, uncentered_major_axes2$cladeplovers, uncentered_major_axes2$phylo)
sandpipers_results <- mapply(vector.diff, uncentered_major_axes2$cladesandpipers, uncentered_major_axes2$phylo)

```{r, fig.height = 18, fig.heigh = 6}
## Plot the blob wise metrics
par(mfrow = c(3, 1))
boxplot(angles, col = colour_vector, ylab = "angle")
boxplot(projections, col = colour_vector, ylab = "projection (scaled)")
boxplot(rejections, col = colour_vector, ylab = "rejection (scaled)")
```

```{r, fig.height = 18, fig.heigh = 6}
## Plot the blob wise metrics
par(mfrow = c(3, 1))
boxplot(angles %% 90, col = colour_vector, ylab = "angle (modulo 90)")
boxplot(abs(projections), col = colour_vector, ylab = "projection (scaled - absolute)")
boxplot(abs(rejections), col = colour_vector, ylab = "rejection (scaled - absolute)")
```

@@

```{r}
library(dispRity)
## Wrapping function
lapply.projections <- function(one_axis, trait_space) {
    return(list(
        projection = projections(as.matrix(trait_space), point1 = one_axis[1,], point2 = one_axis[2,]),
        rejection  = projections(as.matrix(trait_space), point1 = one_axis[1,], point2 = one_axis[2,], measure = "distance")))
}
## Plot rejection/projection profiles
plot.proj.rej <- function(point_list, cols, data, main) {
    ## Colour vector
    colour_vector <- cols[morphdat$clade]
    colour_vector <- adjustcolor(colour_vector, alpha.f = 1/10)
    ## lims
    lim <- range(unlist(point_list))
    plot(NULL, xlim = lim, ylim = lim, xlab = "projection", ylab = "rejection", main = main)
    ## Add each points
    plot.points <- function(one_set, colour_vector) {
        points_args <- list()
        points_args$pch <- 19
        points_args$col <- colour_vector
        points_args$x <- one_set$projection
        points_args$y <- one_set$rejection
        do.call(points, points_args)
    }

@@

```{r}
boxplot.proj.rej <- function(point_list, data, cols) {
    ## Get the classifier
    classifier <- data$clade    
    sort.clade.rej <- function(data, classifier) {
        return(data.frame("points" = data$rejection,
                          "classifier" = classifier))
    }
    sort.clade.proj <- function(data, classifier) {
        return(data.frame("points" = data$projection,
                          "classifier" = classifier))
    }
    ## Sort the data per clade
    rejections <- do.call(rbind, lapply(point_list, sort.clade.rej, classifier))
    projections <- do.call(rbind, lapply(point_list, sort.clade.proj, classifier))
    par(mfrow = c(2,1))
    boxplot(points ~ classifier, data = projections, col = cols, main = "Phylo projections\n(projected on the whole phylo effect)")
    boxplot(points ~ classifier, data = rejections, rejections, col = cols, main = "Phylo rejections")
}

@@

```{r}
## The within clades explo/elabo
gulls_results_uncentered <- lapply(uncentered_major_axes$cladegulls, lapply.projections, trait_space = morphdat[which(morphdat$clade == "gulls"), c(1,2,3)])
plover_results_uncentered <- lapply(uncentered_major_axes$cladeplovers, lapply.projections, trait_space = morphdat[which(morphdat$clade == "plovers"), c(1,2,3)])
sandpipers_results_uncentered <- lapply(uncentered_major_axes$cladesandpipers, lapply.projections, trait_space = morphdat[which(morphdat$clade == "sandpipers"), c(1,2,3)])

>>>>

```{r}
## Get all the phylo main axes
level_centred_major_axes <- get.axes(covar_matrices, centre = mean)
uncentered_major_axes <- get.axes(covar_matrices, centre = "intercept")
```

@@

```{r}
plot(morphdat[, c(1,2)], pch = 19, cex = 0.5,
     col = colour_vector[morphdat$clade],
     xlab = "PC1 (90.5%)",
     ylab = "PC1 (6.86%)")
plot.ellipses(covar_matrices, centre = "intercept",
              col = c("grey",colour_vector),
              add = TRUE)
plot.axes(uncentered_major_axes,
         col = c("grey",colour_vector),
         add = TRUE)
```

@@

```{r}
plot.space(morphdat,
           col = colour_vector,
           levels = morphdat$clade,
           xlab = "PC1 (90.5%)",
           ylab = "PC1 (6.86%)")
plot.ellipses(covar_matrices, centre = mean,
              col = c("grey",colour_vector),
              add = TRUE)
plot.axes(level_centred_major_axes,
         col = c("grey",colour_vector),
         add = TRUE)
```

@@

```{r}
plot.space(morphdat,
           col = colour_vector,
           levels = morphdat$clade,
           xlab = "PC1 (90.5%)",
           ylab = "PC1 (6.86%)",
           main = "Major axes of the ellipses\n(aspect ratio = 1)",
           xlim = c(-1.5, 1.5),
           ylim = c(-1.5, 1.5))
plot.ellipses(covar_matrices, centre = mean,
              col = c("grey",colour_vector),
              add = TRUE)
plot.axes(level_centred_major_axes,
         col = c("grey",colour_vector),
         add = TRUE)
```

@@

```{r}
plot.axes(uncentered_major_axes,
         col = c("grey",colour_vector),
         add = FALSE, main = "Major axes (un-centred)")
```

@@

```{r, eval = FALSE}
source("../Functions/vector.diff.R")
## Get all the angles and stuff
gull_results <- mapply(vector.diff, uncentered_major_axes$cladegulls, uncentered_major_axes$phylo)
plovers_results <- mapply(vector.diff, uncentered_major_axes$cladeplovers, uncentered_major_axes$phylo)
sandpipers_results <- mapply(vector.diff, uncentered_major_axes$cladesandpipers, uncentered_major_axes$phylo)

@@

```{r, fig.height = 18, fig.heigh = 6, eval = FALSE}
## Plot the blob wise metrics
par(mfrow = c(3, 1))
boxplot(angles, col = colour_vector, ylab = "angle")
boxplot(projections, col = colour_vector, ylab = "projection (scaled)")
boxplot(rejections, col = colour_vector, ylab = "rejection (scaled)")
```

@@

```{r, fig.height = 18, fig.heigh = 6, eval = FALSE}
## Plot the blob wise metrics
par(mfrow = c(3, 1))
boxplot(angles %% 90, col = colour_vector, ylab = "angle (modulo 90)")
boxplot(abs(projections), col = colour_vector, ylab = "projection (scaled - absolute)")
boxplot(abs(rejections), col = colour_vector, ylab = "rejection (scaled - absolute)")
```

# Methods bits draft

## The model

We run a MCMCglmm with different levels.

## The mini chains

We used a mini chain approach to:

 * accounted for phylogenetic uncertainty [healy.et.al]
 * accounted for global parameter estimates uncertainty (local vs. global peak)
 * greatly speeding the calculations through high parallelisation

We then combine the mini chains

## Exploration and elaboration analyses

For each model, we randomly draw $N$ variance covariance matrices estimates.
We then used them to analyse the exploration/elaboration component at the species level and at the clade level indepently.

We extracted the major axes of each selected covariance matrices using blablbalba

### At the clade level

We then translated each set of major axes so that they align with each other (i.e. translated the clade major axes origin onto the global major axes origin).

We then calculated the angle between the clade major axes and the global major axes as well as their projection and rejections.

### At the species level

Here we only projected the species coordinates on either:

 * the global major axes to measure the global elaboration/exploration
 * the clade major axes to measure the nested elaboration/exploration

@@

```{r, eval = FALSE}
library(dispRity)
## Wrapping function
lapply.projections <- function(one_axis, trait_space) {
    return(list(
        projection = projections(as.matrix(trait_space), point1 = one_axis[1,], point2 = one_axis[2,]),
        rejection  = projections(as.matrix(trait_space), point1 = one_axis[1,], point2 = one_axis[2,], measure = "distance")))
}
## Plot rejection/projection profiles
plot.proj.rej <- function(point_list, cols, data, main) {
    ## Colour vector
    colour_vector <- cols[morphdat$clade]
    colour_vector <- adjustcolor(colour_vector, alpha.f = 1/10)
    ## lims
    lim <- range(unlist(point_list))
    plot(NULL, xlim = lim, ylim = lim, xlab = "projection", ylab = "rejection", main = main)
    ## Add each points
    plot.points <- function(one_set, colour_vector) {
        points_args <- list()
        points_args$pch <- 19
        points_args$col <- colour_vector
        points_args$x <- one_set$projection
        points_args$y <- one_set$rejection
        do.call(points, points_args)
    }

@@

```{r, eval = FALSE}
boxplot.proj.rej <- function(point_list, data, cols) {
    ## Get the classifier
    classifier <- data$clade    
    sort.clade.rej <- function(data, classifier) {
        return(data.frame("points" = data$rejection,
                          "classifier" = classifier))
    }
    sort.clade.proj <- function(data, classifier) {
        return(data.frame("points" = data$projection,
                          "classifier" = classifier))
    }
    ## Sort the data per clade
    rejections <- do.call(rbind, lapply(point_list, sort.clade.rej, classifier))
    projections <- do.call(rbind, lapply(point_list, sort.clade.proj, classifier))
    par(mfrow = c(2,1))
    boxplot(points ~ classifier, data = projections, col = cols, main = "Phylo projections\n(projected on the whole phylo effect)")
    boxplot(points ~ classifier, data = rejections, rejections, col = cols, main = "Phylo rejections")
}

@@

```{r, eval = FALSE}
## The within clades explo/elabo
gulls_results_uncentered <- lapply(uncentered_major_axes$cladegulls, lapply.projections, trait_space = morphdat[which(morphdat$clade == "gulls"), c(1,2,3)])
plover_results_uncentered <- lapply(uncentered_major_axes$cladeplovers, lapply.projections, trait_space = morphdat[which(morphdat$clade == "plovers"), c(1,2,3)])
sandpipers_results_uncentered <- lapply(uncentered_major_axes$cladesandpipers, lapply.projections, trait_space = morphdat[which(morphdat$clade == "sandpipers"), c(1,2,3)])
